---
title: "Filtered Fit: FCA and the GDSTM"
output: html_notebook
---

## Filtered ML fit and the GDSTM with FRESA.CAD

Here we make use of the **FRESA.CAD::filteredfit()** function to train ML models with and without GDSTM on the ARCENE data set.

> Isabelle Guyon, Steve R. Gunn, Asa Ben-Hur, Gideon Dror, 2004. Result analysis of the NIPS 2003 feature selection challenge. In: NIPS. [\[Web Link\]](http://books.nips.cc/papers/files/nips17/NIPS2004_0194.pdf). *from: <https://archive.ics.uci.edu/ml/datasets/Arcene>*
>
> *a. Original owners\
> The data were obtained from two sources: The National Cancer Institute (NCI) and the Eastern Virginia Medical School (EVMS). All the data consist of mass-spectra obtained with the SELDI technique. The samples include patients with cancer (ovarian or prostate cancer), and healthy or control patients.\
> \
> b. Donor of database\
> This version of the database was prepared for the NIPS 2003 variable and feature selection benchmark by Isabelle Guyon, 955 Creston Road, Berkeley, CA 94708, USA ([isabelle **'\@'** clopinet.com]{.ul}).*

This scrip uses FRESA.CAD and mlbench R packages:

```{r functions,echo = TRUE }
knitr::opts_chunk$set(collapse = TRUE, warning = FALSE, message = FALSE,comment = "#>")

library("FRESA.CAD")
library(mlbench)

op <- par(no.readonly = TRUE)

```

Loading the ARCENE data set

```{r}
trainLabeled <- read.delim("~/GitHub/FCA/Data/trainSet.txt")
validLabeled <- read.delim("~/GitHub/FCA/Data/arcene_valid.txt")
wholeArceneSet <- rbind(trainLabeled,validLabeled)


wholeArceneSet$Labels <-  1*(wholeArceneSet$Labels > 0)
wholeArceneSet[,1:ncol(trainLabeled)] <- sapply(wholeArceneSet,as.double)

table(wholeArceneSet$Labels)

```

Setting some variables for downstream analysis

```{r}
studyName = "Arcene"
datasetframe <- wholeArceneSet
Outcome <- "Labels"

trainFraction = 0.5

```

Setting the Training and Testing sets

```{r, results = "asis", warning = FALSE, dpi=600, fig.height= 6.0, fig.width= 8.0}

tb <- table(datasetframe[,Outcome])
classNames <- unique(datasetframe[,Outcome])

allrowClass <- datasetframe[,Outcome]
names(allrowClass) <- rownames(datasetframe)

trainsize <- trainFraction*min(tb);
trainSamples <- NULL;
for (theClass in classNames)
{
  classSample <- allrowClass[allrowClass == theClass]
  trainSamples <- c(trainSamples,names(classSample[sample(length(classSample),trainsize)]))
}


datasetframe_train <- datasetframe[trainSamples,]
testSamples <- !(rownames(datasetframe) %in% trainSamples)
datasetframe_test <- datasetframe[testSamples,]

outcomes <- datasetframe_train[,Outcome]

pander::pander(table(datasetframe[,Outcome]),caption="All")
pander::pander(table(datasetframe_train[,Outcome]),caption="Training")
pander::pander(table(datasetframe_test[,Outcome]),caption="Testing")


```

## Machine Learning with the filterfit() function

Train a simple Logistic model with LASSO

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 6.0, fig.width= 8.0}


mLASSORaw <- filteredFit(paste(Outcome,"~."),
                   datasetframe_train,
                   fitmethod=LASSO_MIN,
                     filtermethod=univariate_Wilcoxon,
                     filtermethod.control=list(pvalue=0.05,limit= -1),
                    family = "binomial"
                   )


```

Now we run **filterfit**() with decorrelation

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 6.0, fig.width= 8.0}


mLASSODecor <- filteredFit(paste(Outcome,"~."),
                   datasetframe_train,
                    fitmethod=LASSO_MIN,
                     filtermethod=univariate_Wilcoxon,
                     filtermethod.control=list(pvalue=0.05,limit= -1),
                     DECOR = TRUE,
                     family = "binomial"
                   )


```

### Predicting the Testinig set

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 5.0, fig.width= 8.0}


par(mfrow=c(1,2))


psRaw <- predictionStats_binary(cbind(datasetframe_test$Labels,
                                      predict(mLASSORaw,datasetframe_test)),
                                "LASSO Raw",cex=0.75)
pander::pander(psRaw$aucs)


psDecor <- predictionStats_binary(cbind(datasetframe_test$Labels,
                                        predict(mLASSODecor,datasetframe_test)),
                                "LASSO after GDSTM",cex=0.75)
pander::pander(psDecor$aucs)



```
