---
title: "Decorrelation-Based Feature Discovery: Thermal COVID_19"
author: "Jose Tamez"
date: "2022-10-02"
output:
  html_document: 
    toc: yes
    fig_caption: yes
    number_sections: yes
  word_document: 
    reference_docx: WordStyle_FRESA.docx
    toc: yes
    fig_caption: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(collapse = TRUE, warning = FALSE, message = FALSE,comment = "#>")

```

# Effect of GDSTM-Based Decorrelation on Feature Discovery

Here I showcase of to use BSWiMS feature selection/modeling function coupled with Goal Driven Sparse Transformation Matrix (GDSTM) as a pre-processing step to decorrelate highly correlated features. The aim(s) are:

1.  To improve model performance by uncovering the hidden information between correlated features.

2.  To simplify the interpretation of the machine learning models.

This demo will use:

-   *FRESA.CAD::GDSTMDecorrelation()*. For Decorrelation of Multidimensional data sets

    -   *FRESA.CAD::getDerivedCoefficients()*. For the extraction of the model of the newly discovered of decorrelated features.

-   *FRESA.CAD::randomCV()* For the cross-validation of the Machine Learning models

-   *FRESA.CAD::BSWiMS.model()*. For the generation of bootstrapped logistic models

    -   *FRESA.CAD::summary()*. For the summary description of the BSWiMS model

-   *FRESA.CAD::predictionStats_binary()*. For describing the performance of the model

-   *heatmap.2()*. For displaying the correlation matrix

-   *igraph::graph_from_adjacency_matrix()*. For the display of the network of BSWiMS formulas

-   *vioplot::vioplot()*. For the display of the z-distribution of significant features.

### Loading the libraries

```{r}
library("FRESA.CAD")
library(readxl)
library(vioplot)
library(igraph)

op <- par(no.readonly = TRUE)
pander::panderOptions('digits', 3)
pander::panderOptions('table.split.table', 400)
pander::panderOptions('keep.trailing.zeros',TRUE)

```

## Material and Methods

### Signed Log Transform

The function will be used to transform all the continuous features of the data

```{r}
signedlog <- function(x) { return (sign(x)*log(abs(x)+1.0e-12))}

```

## Data: The COVID_19 Data-Set

The data to process is described in:



Saliva Testing Dataset





I added a column to the data identifying the repeated experiments.

```{r}

SalivaThermal <- as.data.frame(read_excel("Data/SalivaThermal_Source_Data_2.xlsx"))

rownames(SalivaThermal) <- paste(SalivaThermal$ID,SalivaThermal$RepID,sep="_")

SalivaThermal$ID <- NULL
SalivaThermal_set1 <- subset(SalivaThermal,RepID==1)
SalivaThermal_set1$ID <- NULL
SalivaThermal_set1$Ct <- NULL
SalivaThermal_set1$class <- 1*(str_detect(rownames(SalivaThermal_set1),"P"))
##The fraction of samples in the training set
pander::pander(table(SalivaThermal_set1$class))
trainFraction=0.95

```



### Training and Testing Sets

We divided the data into training and testing sets.

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 6.0, fig.width= 8.0}
set.seed(2)
caseSet <- subset(SalivaThermal_set1, class == 1)
controlSet <- subset(SalivaThermal_set1, class == 0)
caseTrainSize <- nrow(caseSet)*trainFraction;
controlTrainSize <- nrow(controlSet)*trainFraction;
sampleCaseTrain <- sample(nrow(caseSet),caseTrainSize)
sampleControlTrain <- sample(nrow(controlSet),controlTrainSize)
trainSet <- rbind(caseSet[sampleCaseTrain,], controlSet[sampleControlTrain,])
testSet <-  rbind(caseSet[-sampleCaseTrain,],controlSet[-sampleControlTrain,])
pander::pander(table(trainSet$class))
pander::pander(table(testSet$class))


```

#### Decorrelation: Training and Testing Sets Creation

I compute a decorrelated version of the training and testing sets using the *GDSTMDecorrelation()* function of FRESA.CAD. The first decorrelation will be driven by features associated with the outcome. The second decorrelation will find the GDSTM without the outcome restriction.

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 6.0, fig.width= 8.0}
## The GDSTM transformation driven by the Outcome
deTrain <- GDSTMDecorrelation(trainSet,Outcome="class",thr=0.8,verbose = TRUE)
deTest <- predictDecorrelate(deTrain,testSet)

## The GDSTM transformation without outcome
deTrainU <- GDSTMDecorrelation(trainSet,thr=0.8,verbose = TRUE)
deTestU <- predictDecorrelate(deTrainU,testSet)

```

#### Correlation Matrix of the Decorrelated Test Data

The heat map of the testing set.

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 6.0, fig.width= 8.0}
cormat <- cor(deTest,method="spearman")
gplots::heatmap.2(abs(cormat),
                  trace = "none",
                  scale = "none",
                  mar = c(10,10),
                  col=rev(heat.colors(5)),
                  main = "Test Set Correlation after GDSTM",
                  cexRow = 0.35,
                  cexCol = 0.35,
                  key.title=NA,
                  key.xlab="Spearman Correlation",
                  xlab="Feature", ylab="Feature")

```

### Holdout Cross-Validation

Before doing the feature analysis. I'll explore BSWiMS modeling using the Holdout cross validation method of FRESA.CAD. The purpose of the cross-validation is to observe and estimate the performance gain of decorrelation.

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 3.0, fig.width= 9.0}
par(op)
par(mfrow=c(1,3))

## The Raw validation
cvBSWiMSRaw <- randomCV(SalivaThermal_set1,
                "class",
                fittingFunction= BSWiMS.model,
                classSamplingType = "Pro",
                trainFraction = trainFraction,
                repetitions = 10
)

bpraw <- predictionStats_binary(cvBSWiMSRaw$medianTest,"BSWiMS RAW",cex=0.60)
pander::pander(bpraw$CM.analysis$tab)
pander::pander(bpraw$accc)
pander::pander(bpraw$aucs)
pander::pander(bpraw$berror)

## The validation with Outcome-driven Decorrelation
cvBSWiMSDeCor <- randomCV(SalivaThermal_set1,
                "class",
                trainSampleSets= cvBSWiMSRaw$trainSamplesSets,
                fittingFunction= filteredFit,
                fitmethod=BSWiMS.model,
                filtermethod=NULL,
                DECOR = TRUE,
                DECOR.control=list(Outcome="class",thr=0.8)
)

bpDecor <- predictionStats_binary(cvBSWiMSDeCor$medianTest,"BSWiMS Outcome-Driven GDSTM",cex=0.60)
pander::pander(bpDecor$CM.analysis$tab)
pander::pander(bpDecor$accc)
pander::pander(bpDecor$aucs)
pander::pander(bpDecor$berror)

### Here we compute the probability that the outcome-driven decorrelation ROC is superior to the RAW ROC. 
pander::pander(roc.test(bpDecor$ROC.analysis$roc.predictor,bpraw$ROC.analysis$roc.predictor,alternative = "greater"))

### Testing improving proability
iprob <- .Call("improveProbCpp",cvBSWiMSRaw$medianTest[,2],
               cvBSWiMSDeCor$medianTest[,2],
               cvBSWiMSRaw$medianTest[,1]);
pander::pander(iprob)
### Testing improving accuracy
testRaw <- (cvBSWiMSRaw$medianTest[,1]-cvBSWiMSRaw$medianTest[,2])<0.5
testDecor <- (cvBSWiMSDeCor$medianTest[,1]-cvBSWiMSDeCor$medianTest[,2])<0.5
pander::pander(mcnemar.test(testRaw,testDecor))

## The validation of Decorrelation without the outcome restriction
cvBSWiMSDeCorU <- randomCV(SalivaThermal_set1,
                "class",
                trainSampleSets= cvBSWiMSRaw$trainSamplesSets,
                fittingFunction= filteredFit,
                fitmethod=BSWiMS.model,
                filtermethod=NULL,
                DECOR = TRUE,
                DECOR.control=list(thr=0.8)
)

bpDecorU <- predictionStats_binary(cvBSWiMSDeCorU$medianTest,"BSWiMS Data Driven GDSTM",cex=0.60)
pander::pander(bpDecorU$CM.analysis$tab)
pander::pander(bpDecorU$accc)
pander::pander(bpDecorU$aucs)
pander::pander(bpDecorU$berror)

### Here we compute the probability that the blind decorrelation ROC is superior to the RAW ROC. 

pander::pander(roc.test(bpDecorU$ROC.analysis$roc.predictor,bpraw$ROC.analysis$roc.predictor,alternative = "greater"))
par(op)

## Testing probability improvement
iprob <- .Call("improveProbCpp",cvBSWiMSRaw$medianTest[,2],cvBSWiMSDeCorU$medianTest[,2],cvBSWiMSRaw$medianTest[,1]);
pander::pander(iprob)

## Testing accuracy improvement
testDecorU <- (cvBSWiMSDeCorU$medianTest[,1]-cvBSWiMSDeCorU$medianTest[,2])<0.5
pander::pander(mcnemar.test(testRaw,testDecorU))


```

## The Raw Model *vs.* the Decorrelated-Based Model

After demonstrating that decorrelation is able to improve BSWiMS model performance, I'll focus is showcasing the ability to discover new features associated with the outcome.

First, I'll compute the BSWiMS models for the original data, and for the decorrelated data-set. The model estimation will be done using the training set and tested on the holdout test set, and repeated 10 times. After that, I'll compare the statistical difference of both ROC curves.

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 3.0, fig.width= 9.0}
par(op)
par(mfrow=c(1,3))

bm <- BSWiMS.model(class~.,trainSet,NumberofRepeats = 5)
bpraw <- predictionStats_binary(cbind(testSet$class,predict(bm,testSet)),"BSWiMS RAW",cex=0.60)

bmd <- BSWiMS.model(class~.,deTrain,NumberofRepeats = 5)
bpdecor <- predictionStats_binary(cbind(deTest$class,predict(bmd,deTest)),"Outcome-Driven Decorrelation",cex=0.60)

## Comparing the two ROC curves
pander::pander(roc.test(bpdecor$ROC.analysis$roc.predictor,bpraw$ROC.analysis$roc.predictor,alternative = "greater"))
## Comparing the test accuracy
testRaw <- (testSet$class-predict(bm,testSet))<0.5
testDecor <- (deTest$class-predict(bmd,deTest))<0.5
pander::pander(mcnemar.test(testRaw,testDecor))


bmdU <- BSWiMS.model(class~.,deTrainU,NumberofRepeats = 5)
bpdecorU <- predictionStats_binary(cbind(deTest$class,predict(bmdU,deTestU)),"Blind Decorrelation",cex=0.60)

## Comparing the test curves
pander::pander(roc.test(bpdecorU$ROC.analysis$roc.predictor,bpraw$ROC.analysis$roc.predictor,alternative = "greater"))
## Comparing the accuracy
testDecorU <- (deTestU$class-predict(bmdU,deTestU))<0.5
pander::pander(mcnemar.test(testRaw,testDecorU))

par(op)

```

## The Feature Associations

I'll print the graph showing the association between features. Each feature cluster represents a logistic regression formula (formula nugget) discovered by the BSWiMS method. The figure will plot:

-   Raw formula network

-   Outcome-driven network

-   Blind network

The plots will show only formula networks with more than 50% of occurrence and 25% of feature to feature association.

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 3.0, fig.width= 9.0}
par(op)
par(mfrow=c(1,3))
### The raw model

pander::pander(nrow(bm$bagging$formulaNetwork))


cmax <- apply(bm$bagging$formulaNetwork,2,max)
cnames <- names(cmax[cmax>=0.5])
cmax <- cmax[cmax>=0.5]
adma <- bm$bagging$formulaNetwork[cnames,cnames]

rownames(adma) <- cnames
colnames(adma) <- cnames
names(cmax) <- cnames
adma[adma<0.25] <- 0;
gr <- graph_from_adjacency_matrix(adma,mode = "undirected",diag = FALSE,weighted=TRUE)
gr$layout <- layout_with_fr

fc <- cluster_optimal(gr)
plot(fc, gr,
     vertex.size=20*cmax,
     vertex.label.cex=0.5,
     vertex.label.dist=0,
     main="Original Feature Association")



### The Outcome Driven Model

pander::pander(nrow(bmd$bagging$formulaNetwork))


cmax <- apply(bmd$bagging$formulaNetwork,2,max)
cnames <- names(cmax[cmax>=0.5])
outcomeNames <- cnames

cmax <- cmax[cmax>=0.5]
adma <- bmd$bagging$formulaNetwork[cnames,cnames]

rownames(adma) <- cnames
colnames(adma) <- cnames
names(cmax) <- cnames
adma[adma<0.25] <- 0;
gr <- graph_from_adjacency_matrix(adma,mode = "undirected",diag = FALSE,weighted=TRUE)
gr$layout <- layout_with_fr

fc <- cluster_optimal(gr)
clusterOutcome <- fc
clusterOutcome$names <- outcomeNames

plot(fc, gr,
     vertex.size=20*cmax,
     vertex.label.cex=0.5,
     vertex.label.dist=0,
     main="Outcome-Driven Decorrelation")


### The Blind Decorrelation

pander::pander(nrow(bmdU$bagging$formulaNetwork))


cmax <- apply(bmdU$bagging$formulaNetwork,2,max)
cnames <- names(cmax[cmax>=0.5])
cmax <- cmax[cmax>=0.5]
adma <- bmdU$bagging$formulaNetwork[cnames,cnames]

rownames(adma) <- cnames
colnames(adma) <- cnames
names(cmax) <- cnames
adma[adma<0.25] <- 0;
gr <- graph_from_adjacency_matrix(adma,mode = "undirected",diag = FALSE,weighted=TRUE)
gr$layout <- layout_with_fr

fc <- cluster_optimal(gr)
plot(fc, gr,
     vertex.size=20*cmax,
     vertex.label.cex=0.5,
     vertex.label.dist=0,
     main="Blind Decorrelation")

```

### Feature Analysis of Models

The analysis of the features required to predict the outcome will use the following:

1.  Analysis of the BSWiMS bagged model using the summary function.

2.  Analysis of the sparse GDSMT

3.  Analysis of the univariate association of the model features of both models

4.  Report the new features not found by the Original data analysis

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 6.0, fig.width= 8.0}
par(op)
par(mfrow=c(1,1))
## 1 Get the Model Features
smOriginal <- summary(bm)
rawnames <- rownames(smOriginal$coefficients)

### From Drived Decorrelation
smDecor <- summary(bmd)
decornames <- rownames(smDecor$coefficients)

### From Blind Decorrelation
smDecorU <- summary(bmdU)
decornamesU <- rownames(smDecorU$coefficients)



## 2 Get the decorrelation matrix formulas
dc <- getDerivedCoefficients(deTrain)
### 2a Get only the ones that were decorrelated by the decorrelation-based model
deNames_in_dc <- decornames[decornames %in% names(dc)]
selectedlist <- dc[deNames_in_dc]
theDeFormulas <- selectedlist
pander::pander(selectedlist)
names(selectedlist) <- NULL
### 2b Get the the names of the original features

allDevar <- unique(c(names(unlist(selectedlist)),decornames))
allDevar <- allDevar[!str_detect(allDevar,"De_")]
allDevar <- str_remove(allDevar,"Ba_")
allDevar <- unique(allDevar)


# The analysis of the blind decorrelation

dcU <- getDerivedCoefficients(deTrainU)
### 2a Get only the ones that were decorrelated by the decorrelation-based model
deNames_in_dcU <- decornamesU[decornamesU %in% names(dcU)]
selectedlistU <- dcU[deNames_in_dcU]
pander::pander(selectedlistU)
names(selectedlistU) <- NULL
### 2b Get the the names of the original features

allDevarU <- unique(c(names(unlist(selectedlistU)),decornamesU))
allDevarU <- allDevarU[!str_detect(allDevarU,"De_")]
allDevarU <- str_remove(allDevarU,"Ba_")
allDevarU <- unique(allDevarU)

pander::pander(c(length(rawnames),length(decornames),length(decornamesU)))
pander::pander(c(length(rawnames),length(allDevar),length(allDevarU)))


### 2c Get only the new feautres not found in the original analysis
dvar <- allDevar[!(allDevar %in% rawnames)] 

### 2d Get the decorrelated variables that have new features
newvars <- character();
for (cvar in deNames_in_dc)
{
  lvar <- dc[cvar]
  names(lvar) <- NULL
  lvar <- names(unlist(lvar))
  if (length(lvar[lvar %in% dvar]) > 0)
  {
     newvars <- append(newvars,cvar)
  }
}

## 3 Here is the univariate z values of the orignal set
#pander::pander(bm$univariate[dvar,])
## 4 Here is the univariate z values of the decorrelated set
#pander::pander(bmd$univariate[newvars,])

## 4a The scater plot of the decorrelated vs original Univariate values

zvalueNew <- bmd$univariate[newvars,]
rownames(zvalueNew) <- str_remove(rownames(zvalueNew),"De_")
rownames(zvalueNew) <- str_remove(rownames(zvalueNew),"Ba_")

zvaluePrePost <- bm$univariate[rownames(zvalueNew),c(1,3)]
zvaluePrePost$Name <- NULL
zvaluePrePost$NewZ <- zvalueNew[rownames(zvaluePrePost),"ZUni"]
pander::pander(zvaluePrePost)
plot(zvaluePrePost,
     xlim=c(-0.5,6.5),
     ylim=c(0,7),
     xlab="Original Z",
     ylab="Decorrelated Z",
     main="Unviariate IDI Z Values",
     pch=3,cex=0.5,
     col="red")
abline(v=1.96,col="blue")
abline(h=1.96,col="blue")
text(zvaluePrePost$ZUni,zvaluePrePost$NewZ,rownames(zvaluePrePost),srt=65,cex=0.75)


```

### The Summary of the Decorrelated-Based Model

Here I will print the summary statistics of the Logistic models found by BSWiMS, using the original and transformed dataset. After that, I will show the characteristics of the features not found by the original analysis.

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 6.0, fig.width= 8.0}

pander::pander(smOriginal$coefficients)

pander::pander(smDecor$coefficients)

pander::pander(smDecorU$coefficients)

## Let focus on the new features

decorCoeff <- smDecor$coefficients[newvars,];
ncoef <- dc[newvars]
cnames <- lapply(ncoef,names)
names(cnames) <- NULL;
decorCoeff$Elements <- lapply(cnames,paste,collapse="+")
pander::pander(decorCoeff)
```

## Differences Between Blind *vs.* Outcome-Driven Decorrelation

In this section I will show the differences in unaltered basis vectors between the Outcome driven Transformation *vs.* the blind decorrelated transformation

```{r results = "asis", warning = FALSE, dpi=600, fig.height= 6.0, fig.width= 8.0}
par(op)
par(mfrow=c(1,1))


smDecorU <- summary(bmdU)
decornamesU <- rownames(smDecorU$coefficients)

get_De_names <- decornames[!str_detect(decornames,"De_")]
get_De_namesU <- decornamesU[!str_detect(decornamesU,"De_")]

unn <- bmd$univariate[,3]
names(unn) <- rownames(bmd$univariate)
pander::pander(as.matrix(unn[get_De_names]))
pander::pander(summary(unn[get_De_names]))

unnU <- bmdU$univariate[,3]
names(unnU) <- rownames(bmdU$univariate)
pander::pander(as.matrix(unnU[get_De_namesU]))
pander::pander(summary(unnU[get_De_namesU]))
#boxplot(unn[get_De_names],unnU[get_De_namesU],xlab=c("Method"),ylab="Z",main="Z Values of Basis Features")

x1 <- unn[get_De_names]
x2 <- unnU[get_De_namesU]
X3 <- x1[!(get_De_names %in% get_De_namesU)]
X4 <- x2[!(get_De_namesU %in% get_De_names)]
vioplot(x1, x2, X3,X4, 
        names = c("Outcome-Driven", 
                  "Blind",
                  "Not in Blind",
                  "Not in Outcome-Driven"),
        ylab="Z IDI",
   col="gold")
title("Violin Plots of Unaltered-Basis")

sameFeatures <- get_De_names[get_De_names %in% get_De_namesU]
pander::pander(as.matrix(unn[sameFeatures]))
## The features by Outcome Drive not in Blind
pander::pander(as.matrix(x1[!(get_De_names %in% get_De_namesU)]))

## The features not in outcome driven
pander::pander(as.matrix(x2[!(get_De_namesU %in% get_De_names)]))
```

### The Final Table

I'll create a table subset of the logistic model from the Outcome-Driven decorrelated data.

The table will have:

1.  The top associated features described by the feature network, as well as, and the new features.

    1.  For Decorrelated features it will provide the decorrelation formula

2.  Nugget labels

    1.  The label of nugget as found by the clustering procedure

3.  The feature coefficient

4.  The feature Odd ratios and their corresponding 95%CI

```{r}

## The features in top nugget
clusterFeatures <- clusterOutcome$names
## The new features 
discoveredFeatures <- newvars[zvaluePrePost$ZUni<1.96]

tablefinal <- smDecor$coefficients[unique(c(clusterFeatures,discoveredFeatures)),
                                   c("Estimate","lower","OR","upper","z.IDI")]

nugget <- clusterOutcome$membership
names(nugget) <- clusterOutcome$names
tablefinal$Nugget <- nugget[rownames(tablefinal)]
tablefinal$Nugget[is.na(tablefinal$Nugget)] <- "D"
deFromula <- character(length(theDeFormulas))
names(deFromula) <- names(theDeFormulas)
for (dx in names(deFromula))
{
  coef <- theDeFormulas[[dx]]
  cname <- names(theDeFormulas[[dx]])
  names(cname) <- cname
  for (cf in names(coef))
  {
    if (cf != dx)
    {
      if (coef[cf]>0)
      {
        deFromula[dx] <- paste(deFromula[dx],
                               sprintf("+ %5.3f*%s",coef[cf],cname[cf]))
      }
      else
      {
        deFromula[dx] <- paste(deFromula[dx],
                               sprintf("%5.3f*%s",coef[cf],cname[cf]))
      }
    }
  }
}
tablefinal$DecorFormula <- deFromula[rownames(tablefinal)]
pander::pander(tablefinal)

```

#### Saving all the generated data

```{r}
save.image("~/GitHub/FCA/COVID_19Demo.RData")

```
